### 2013-09-26 manuscripts/dpp-msbayes/dpp-msbayes.tex
Finishing draft of dpp model and ABC description.
Starting on simulation methods.

@@

\subsection*{ABC estimation of the posterior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameters values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{}, from their respective
prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots, \divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequences alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments (these are the same statistics calculated from the observed
alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statstic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random sample
of parameters
from the model prior and there associated summary statistics
.

This has the advantage of reducing the sample space of the number of
divergence models, but has two disadvantages.
First, additional information about the data is lost.
By throwing away the identity of the \npairs{} pairs of populations, all
information associating each pair with its amount of data (e.g., the number gene copies
collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci),
information content in the data (i.e., the amount of sequence variation), as
well as its taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
The original descriptions of the \msb model claim that the independence
between the sample sizes \popSampleSize{1}{}, \popSampleSize{2}{} and
the summary statistic $\pi_b$ calculated from the alignment for each
population pair allows exchangeability.
This is incorrect.
The other summary statistics, and the alignments they are based on are
not independent of the sample sizes.
The alignments and summary statitics are not identically distributed because of
taxon- and locus-specific parameters \hkyModel{}{},
\mutationRateScalarConstant{}{}, and \ploidyScalar{}{}
Second, the identity of the pairs of populations is lost.

@@

\renewcommand{\refname}{\noindent\MakeUppercase{\LARGE\sffamily\upshape supporting information}}

>>>>

\subsubsection*{Estimating ordered divergence models}
As mentioned above, \msb samples over unordered divergence models
(i.e., unordered partitions of the \npairs{} pairs of populations).
That is, the identity of each population pair, and all the information
associated with it, is discarded.
In our implementation, inference can be done on either unordered or ordered
models of divergence.
This is discussed in more detail in the description of the ABC implementation
below.

@@

\subsection*{ABC estimation of the posterior}
\subsubsection*{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameters values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{},
from their respective prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequences alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments (these are the same statistics
calculated from the observed alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statistic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and there associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

\subsubsection*{Ordering of taxon-specific summary statistics}
As eluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, they, along with the observed summary statistics, are re-ordered by
descending values of average pairwise differences between the descendant
populations \citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of discrete
divergence models \divTimeMapVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer directly associated to the taxa under study.

The original descriptions of the \msb model claim that this re-ordering is
justified by the fact that of $\pi_b$ (and other summary statistics) are
unrelated to the sample sizes \popSampleSize{1}{} and \popSampleSize{2}{} of
each pair and are thus exchangeable \citet{Hickerson2006,Huang2011}.
This is actually incorrect for two of reasons.
First, the expectation of $\pi_b$ is not independent of samples sizes.
If there are more than one coalescent events in the ancestor, which is expected
to be common on phylogeographic timescales, more samples will increase the
probability of capturing these deeper events, and thus affect the average
pairwise differences between the descendant populations.
Also, other statistics that estimate gross diversity (e.g., $\pi$ and
$\theta_W$) are clearly not independent of sample size.
Second, for variables to be exchangeable, they do not need to be independent,
but their marginal distributions must be the same (i.e., they must be
identically distributed).
The simulated alignments and their summary statistics are not identically
distributed because of differing sample sizes \emph{and} taxon- and
locus-specific parameters \hkyModel{}{}, \mutationRateScalarConstant{}{}, and
\ploidyScalar{}{}.

Thus, the theoretical basis for this reshuffling is questionable.
It may be justifiable as an additional coarsening of the data,
by removing the information associated with the identity of the
population pairs.
However, we do not provide a formal proof that this re-ordering of the pairs
does not introduce bias.
One can imagine situations in which the sampling intensity (i.e., the
number of gene copies, loci, and locus length) is highly skewed across
pairs and/or there is large heterogeneity in mutational processes
(e.g., \hkyModel{}{}, \mutationRateScalarConstant{}{}) among the pairs.
In such cases it seems possible that discarding this information could
cause bias.
Furthermore, given that part of the motivation for re-ordering by $\pi_b$ is to
minimize Euclidean distances between simulated datasets in which the true model
has a single divergence event, and this reordering increases the model's
tendency to infer a single divergence event \citep{Huang2011}, it seems
possible that this approach could be biasing the method \citet{Oaks2012}.

To maintain compatibility and comparability with \msb we maintain
the re-ordering of taxon-specific summary statistics by $\pi_b$, but
we also allow the order to be maintained.
This allows the estimation of ordered divergence models.

\subsubsection*{Obtaining an approximate posterior from the prior samples}
We use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and then
it is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
the \ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the number of desired
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.

\subsection*{Simulation-based validation of the model}

@@

\renewcommand{\refname}{\noindent\MakeUppercase{\LARGE\sffamily\upshape Supporting Information}}
